---
title: 'Pain Management in The ICU Longitudinal Study: Statistical Report'
author: "Dr. Benjamin Moran"
output:
  word_document: default
  pdf_document: default
  html_document: 
    toc: true
    toc_float: true
editor_options: 
  markdown: 
    wrap: 72
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)
```


## Part 1: Methods and Analysis

[**Aims/Objectives**]{.ul}

1.  Describe the pain assessment tools and frequency used in ICU
    patients.
2.  Describe the opioid management of ICU patients.
3.  Describe the adjuvant analgesia used in ICU patients.
4.  Describe the Trajectory of Pain Scores and Opioid Dosing in Mechanically Ventilated ICU Patients.
5.  Describe the Trajectory of Sedation Scores and Sedation Dosing in MV ICU Patients.
6.  Describe the Trajectory of Analgesia and Sedation Delivery over a 24 hour period.


[**Methods**]{.ul}

This manuscript has been prepared in accordance with the Strengthening
the Reporting of Observational Studies in Epidemiology (STROBE)
Statement [@vonelm2008].

[**Statistical Analysis**]{.ul}

Descriptive statistics were used for demographic and clinical data.
Normally distributed variables were reported as means (standard
deviations) and non-normally distributed variables as medians
(interquartile ranges). Proportions were reported as percentages.
No assumptions were made about missing data. Statistical analysis
was performed using R Version 4.1.0 and RStudio Version 1.4.1714
(RStudio Team (2021). RStudio: Integrated Development for R. RStudio,
PBC, Boston, MA URL <http://www.rstudio.com/>) statistical software.
Packages used for analysis included tidyverse, ggplot2, gtsummary and gt.
*p* values \<0.05 were considered significant.\


\
\
\
## Part 2: Patient Demographics

```{r echo=FALSE, fig.align='center', message=FALSE, warning=FALSE}

library(tidyverse)
library(gtsummary)
# Open and attach CSV file #
demogr <- readxl::read_excel("Data/Test CCLHD_Gos 01_01_22 to 30_03_22 Version_2.xlsx")

# Create Age Variable from DOB and ICU Admission Date
demogr <- demogr %>%
  mutate(age = lubridate::as.duration(`Admission Date` - PatientDOB) |>  as.numeric('years'))

# Create ICU LOS Variable
demogr <- demogr %>%
  mutate(icu_los = lubridate::as.duration(`ICU Discharge Date/Time` - `Admission Date`) |>  as.numeric('days'))

# Create Hospital LOS Variable
demogr <- demogr %>%
  mutate(hosp_los = lubridate::as.duration(`Hospital Discharge Date/Time` - HospAD) |>  as.numeric('days'))

# Create Variables for admission diagnoses
demogr <- demogr |> 
  mutate(AP3Diag = case_when(
    (AP3Diagnosis...11>=1301) & (AP3Diagnosis...11<=1304) ~ "Post-Operative",
    (AP3Diagnosis...11>=1401) & (AP3Diagnosis...11<=1413) ~ "Post-Operative",
    (AP3Diagnosis...11>=1602) & (AP3Diagnosis...11<=1605) ~ "Post-Operative",
    (AP3Diagnosis...11>=1701) & (AP3Diagnosis...11<=1705) ~ "Post-Operative",
    (AP3Diagnosis...11>=1801) & (AP3Diagnosis...11<=1803) ~ "Post-Operative",
    (AP3Diagnosis...11>=1902) & (AP3Diagnosis...11<=1904) ~ "Post-Operative",
    (AP3Diagnosis...11==2101) ~ "Post-Operative",
    (AP3Diagnosis...11==2201) ~ "Post-Operative",
    (AP3Diagnosis...11>=101) & (AP3Diagnosis...11<=111) ~ "Medical",
    (AP3Diagnosis...11>=201) & (AP3Diagnosis...11<=213) ~ "Medical",
    (AP3Diagnosis...11>=301) & (AP3Diagnosis...11<=313) ~ "Medical",
    (AP3Diagnosis...11>=406) & (AP3Diagnosis...11<407) ~ "Medical",
    (AP3Diagnosis...11>=501) & (AP3Diagnosis...11<=504) ~ "Sepsis",
    (AP3Diagnosis...11>601) & (AP3Diagnosis...11<=605) ~ "Trauma",
    (AP3Diagnosis...11>=701) & (AP3Diagnosis...11<=704) ~ "Medical",
    (AP3Diagnosis...11>=801) & (AP3Diagnosis...11<=802.11) ~ "Medical",
    (AP3Diagnosis...11>=901) & (AP3Diagnosis...11<=903) ~ "Medical",
    (AP3Diagnosis...11>=1101) & (AP3Diagnosis...11<=1102) ~ "Medical",
    (AP3Diagnosis...11 == 212.03 | AP3Diagnosis...11 == 213.02) ~ "COVID"))

# Reorder Apache3 Diagnosis
demogr <- demogr %>% mutate(AP3Diag = factor(AP3Diag, levels = c("Medical", "Sepsis", "Trauma", "Post-Operative", "COVID")))

# Reorder Admission Source
demogr <- demogr %>% mutate(`source of admission` = factor(`source of admission`, levels = c("Emergency department", "OT/Recovery", "Ward", "Other hospital ICU/HDU")))

# Create COVID Binary Variable
demogr <- demogr %>%
  mutate(COVID = as.character(AP3Diagnosis...11 %in% c(212.03, 213.02))) 

# Convert Clinical Frailty Score (CFS) to Categorical
demogr <- demogr %>%
  mutate(CFS = case_when(
    Fraility %in% c("Very Fit", "Well", "Managing Well") ~ "Fit",
    Fraility %in% "Vulnerable" ~ "Vulnerable",
    Fraility %in% c("Mildly Frail", "Moderately Frail", "Severely Frail", "Very Severely Frail", "Terminally Ill") ~ "Frail")) %>% 
  # Reorder Apache3
  mutate(CFS = factor(CFS, levels = c("Fit", "Vulnerable", "Frail")))

# Combine patient data that were reintubated (same ICU admission, but have duplicate rows to account for different ventilation event) 
demogr <- demogr |> 
  group_by(PatientID) |> # This identifier is the same as long as it's the same ICU admission
  mutate(`Vent start time` = list(`Vent start time`), `Vent End time` = list(`Vent End time`)) |>
  unnest_wider(c(`Vent start time`, `Vent End time`), names_sep = "_") |> 
  distinct() |> # removes duplicate rows
  mutate(retube = ifelse(is.na(`Vent start time_2`), 0, 1)) # Create binary variable for whether the pt was reintubated


# Combine patient data that had separate ICU admissions (they have separate PatientID numbers)
demogr <- demogr |> 
  group_by(PatientDOB, PatientSex, HospAD) |> # Only way to group the variables to determine if it's the same patient as the PatientID won't be the same
  mutate(`Admission Date` = list(`Admission Date`), `ICU Discharge Date/Time` = list(`ICU Discharge Date/Time`)) |>
  unnest_wider(c(`Admission Date`, `ICU Discharge Date/Time`), names_sep = "_") |> 
  mutate(readmit = ifelse(is.na(`Admission Date_2`), 0, 1)) # Create binary variable for whether the pt was readmitted

# NOTE- the readmit variable will report the same patient as they haven't been combined.

# Create Duration of Ventilation Variable
demogr <- demogr %>% 
  mutate(vent_dur = lubridate::as.duration(`Vent End time` - `Vent start time`) %>% as.numeric('days'))


# Generate Table 1
demogr %>% 
  select(age, PatientSex, APACHE3, AP3Diag, 'source of admission', vent_dur, CFS, icu_los, `ICU Mortality`, hosp_los, `Hospital Mortality`) %>% 
  tbl_summary(missing = "no",
              label = list(age="Age in years, median (IQR)",
                           PatientSex="Female, number (%)",
                           APACHE3="APACHE III score, median (IQR)",
                           AP3Diag="Admission diagnosis, number (%)",
                           'source of admission' ="Admission source, number (%)",
                           CFS = "Frailty, number (%)",
                           icu_los = "ICU length of stay, median days (IQR)",
                           hosp_los = "Hospital length of stay, median days (IQR)",
                           vent_dur = "Duration of ventilation, median days (IQR)"),
              value = list(PatientSex ~ "Female",
                           'ICU Mortality' ~ "Yes",
                           'Hospital Mortality' ~ "Yes" ),
              statistic = list(all_continuous() ~ "{median} ({p25}-{p75})",
                               all_categorical() ~ "{n} ({p}%)")) %>%  
  bold_labels() %>% 
  italicize_levels() %>% 
  modify_caption("**Table 1. Patient Demographics**") %>% 
  
  modify_table_body(
    ~.x %>% 
      
      # add demographic grouping variable
      rbind(
        tibble(
          variable="Demographics",
          var_type=NA,
          var_label = "Demographics",
          row_type="label",
          label="Demographics",
          stat_0= NA)) %>% 
      
      # add admission grouping variable
      rbind(
        tibble(
          variable="Admission Characteristics",
          var_type=NA,
          var_label = "Admission Characteristics",
          row_type="label",
          label="Admission Characteristics",
          stat_0= NA)) %>%  
      
      # add outcome grouping variable
      rbind(
        tibble(
          variable="Outcome Characteristics",
          var_type=NA,
          var_label = "Outcome Characteristics",
          row_type="label",
          label="Outcome Characteristics",
          stat_0= NA)) %>%  
      
      # specify the position you want these in
      
      arrange(factor(variable, levels=c("Demographics",
                                        "age",
                                        "PatientSex",
                                        "Admission Characteristics",
                                        "APACHE3",
                                        "AP3Diag",
                                        "source of admission",
                                        "CFS",
                                        "Outcome Characteristics",
                                        "vent_dur",
                                        "icu_los",
                                        "ICU Mortality",
                                        "hosp_los",
                                        "Hospital Mortality"))))%>%
  
  # and you can then indent the actual variables
  modify_column_indent(columns=label, rows=variable %in% c("age",
                                                           "PatientSex",
                                                           "APACHE3",
                                                           "AP3Diag",
                                                           "source of admission",
                                                           "CFS",
                                                           "vent_dur",
                                                           "icu_los",
                                                           "ICU Mortality",
                                                           "hosp_los",
                                                           "Hospital Mortality"))%>%
  
  # and double indent their levels
  modify_column_indent(columns=label, rows= (variable %in% c("AP3Diag",
                                                             "source of admission",
                                                             "CFS") 
                                             & row_type=="level"),
                       double_indent=T)

```
\
\
\
## Part 3: Daily Pain Scores
```{r echo=FALSE, fig.align='center', message=FALSE, warning=FALSE}

library(janitor)
library(tidyverse)
library(lubridate)
library(openxlsx)

# Import data
pain <- read_excel("2. Daily Pain Scores.xlsx")

# Convert Time integer into Datetime
pain <- pain |> mutate(Time = convertToDateTime(pain$Time, origin = "1900-01-01"))

#Clean Column Names
pain <- pain |>  
  clean_names()

# Clean Up Data Frame
pain <- pain |>  
  filter(ventday != "NULL", assessment_tool !="NPS_static", assessment_tool != "NPS_dyn") |> 
  mutate(ventday = as.numeric(ventday))

# Create Time Difference Variable Between Pain Scores
pain <- pain |> 
  group_by(patient_id) |> 
  complete(time = seq.POSIXt(min(ceiling_date(time , 'day')), 
                             max(floor_date(time , 'day')), by = 'day')) |> # Generates a time point at midnight
  arrange(patient_id, time) |> 
  fill(c(patient_id, assessment_tool, score)) |> fill(ventday, .direction = "up") |> 
  mutate(time_diff_hours = as.numeric(lead(time) - time, units = 'hours')) 

# Change a ventialtion day that was erroneous
# Need to find 
pain <- pain |> 
  mutate(ventday = case_when(
    patient_id == 20849 & time == "2022-01-29 00:00:00" ~ 1,
    TRUE ~ ventday
  ))

# Create pain categories
pain <- pain %>% 
  mutate(pain_cat = case_when(
    # CPOT Assessment Tool
    (assessment_tool == "CPOT") & (score == 0) ~ "None",
    (assessment_tool == "CPOT") & (score >= 1) & (score <=2) ~ "Mild",
    (assessment_tool == "CPOT") & (score >= 3) & (score <=5) ~ "Moderate",
    (assessment_tool == "CPOT") & (score >= 6) & (score <=8) ~ "Severe",
    # NPS_static Assessment Tool
    (assessment_tool == "NPS_static") & (score == 0) ~ "None",
    (assessment_tool == "NPS_static") & (score >= 1) & (score <=3) ~ "Mild",
    (assessment_tool == "NPS_static") & (score >= 4) & (score <=7) ~ "Moderate",
    (assessment_tool == "NPS_static") & (score >= 8) & (score <=10) ~ "Severe",
    # NPS_dyn Assessment Tool
    (assessment_tool == "NPS_dyn") & (score == 0) ~ "None",
    (assessment_tool == "NPS_dyn") & (score >= 1) & (score <=3) ~ "Mild",
    (assessment_tool == "NPS_dyn") & (score >= 4) & (score <=7) ~ "Moderate",
    (assessment_tool == "NPS_dyn") & (score >= 8) & (score <=10) ~ "Severe",
    # BPS Assessment Tool
    (assessment_tool == "BPS") & (score <= 3) ~ "None",
    (assessment_tool == "BPS") & (score >= 4) & (score <=6) ~ "Mild",
    (assessment_tool == "BPS") & (score >= 7) & (score <=9) ~ "Moderate",
    (assessment_tool == "BPS") & (score >= 10) & (score <=12) ~ "Severe"
  ))

# Create Time in No Pain
pain <- pain %>% 
  mutate(pain_none = ifelse(pain_cat == "None", time_diff_hours, 0))

# Create Time in Mild Pain
pain <- pain %>% 
  mutate(pain_mild = ifelse(pain_cat == "Mild", time_diff_hours, 0))

# Create Time in Moderate Pain
pain <- pain %>% 
  mutate(pain_mod = ifelse(pain_cat == "Moderate", time_diff_hours, 0))

# Create Time in Severe Pain
pain <- pain %>% 
  mutate(pain_sev = ifelse(pain_cat == "Severe", time_diff_hours, 0))

# Create Time in Mild to Severe Pain (ie any pain)
pain <- pain |> 
  mutate(pain_mild_sev = pain_mild + pain_mod + pain_sev)

# Create Time in Moderate to Severe Pain
pain <- pain %>% 
  mutate(pain_mod_sev = pain_mod + pain_sev)

# Convert NAs into 0
pain <- pain |> 
  mutate(across(where(is.numeric), replace_na, 0))

# Create Variable of % Time in Each Pain Category Per Day (Excluding Ventday = NULL and NPS scores as they usually are done with CPOT in ventilated patients)
pain <- pain |> 
  select(patient_id, logical_unit_id, ventday, pain_none, pain_mild, pain_mod, pain_sev, pain_mild_sev, pain_mod_sev) |> 
  mutate(across(where(is.numeric), round, 1)) |> 
  group_by(patient_id, ventday) |>  
  summarise(across(pain_none: pain_mod_sev, sum))

# Check Number of Hours
pain <- pain |> 
  mutate(total_pain_hrs = pain_none + pain_mild + pain_mod + pain_sev)

# Remove numbers at end of variable names
new_var_name <- c(pain_none = "pain_none_1", pain_mild = "pain_mild_1", pain_mod = "pain_mod_1", pain_sev = "pain_sev_1", pain_mod_sev = "pain_mod_sev_1")
pain <- pain %>% 
  rename(all_of(new_var_name)) |> 
  arrange(patient_id, ventday)


# Convert to wide format
pain <- pain %>% 
  pivot_wider(
    id_cols = patient_id,
    names_from = c(ventday),
    names_glue = "vent{ventday}_{.value}",
    values_from = c(pain_none, pain_mild, pain_mod, pain_sev, pain_mod_sev),
    values_fill = 0)

```
\
\
\
## Part 4: Daily Opioid Consumption

```{r echo=FALSE, fig.align='center', message=FALSE, warning=FALSE}

opioid <- read_excel("3. Daily Opioids.xlsx")
library(janitor)
library(tidyverse)
library(lubridate)
library(zoo)

#Clean Column Names
opioid <- opioid %>% 
  clean_names()

# Rename variables naming opioid infusions and dosage time intervals
opioid_var <- c(morphinf = "medication_5", fentinf = "medication_14", hydroinf = "medication_23",
                morph0_3 = "x00_03_6", morph3_6 = "x03_06_7", morph6_9 = "x06_09_8", morph9_12 = "x09_12_9", morph12_15 = "x12_15_10", morph15_18 = "x15_18_11", morph18_21 = "x18_21_12", morph21_24 = "x21_00_13",
                fent0_3 = "d00_03", fent3_6 = "d03_06", fent6_9 = "d06_09", fent9_12 = "d09_12", fent12_15 = "d12_15", fent15_18 = "d15_18", fent18_21 = "d18_21", fent21_24 = "d21_00",
                hydro0_3 = "x00_03_24", hydro3_6 = "x03_06_25", hydro6_9 = "x06_09_26", hydro9_12 = "x09_12_27", hydro12_15 = "x12_15_28", hydro15_18 = "x15_18_29", hydro18_21 = "x18_21_30", hydro21_24 = "x21_00_31")

opioid <- opioid %>% 
  rename(all_of(opioid_var))

# Change character to numeric
opioid <- opioid %>% 
  mutate(across(c(`morph0_3`:`morph21_24`, `fent0_3`:`fent21_24`, `hydro0_3`:`hydro21_24`), as.numeric))
 
# Change NA to 0
opioid <- opioid %>% 
  replace(is.na(.), 0)

# Create Time Epochs OME
opioid <- opioid %>% 
  rowwise() %>% 
  mutate(
    ome0_3 = morph0_3*3 + fent0_3*2 + hydro0_3*3,
    ome3_6 = morph3_6*3 + fent3_6*2 + hydro3_6*3,
    ome6_9 = morph6_9*3 + fent6_9*2 + hydro6_9*3,
    ome9_12 = morph9_12*3 + fent9_12*2 + hydro9_12*3,
    ome12_15 = morph12_15*3 + fent12_15*2 + hydro12_15*3,
    ome15_18 = morph15_18*3 + fent15_18*2 + hydro15_18*3,
    ome18_21 = morph18_21*3 + fent18_21*2 + hydro18_21*3,
    ome21_24 = morph21_24*3 + fent21_24*2 + hydro21_24*3
    ) %>%
  ungroup()

# Graph Opioid OME 
opioid_long <- opioid %>% 
  pivot_longer(c(`ome0_3`, `ome3_6`, `ome6_9`, `ome9_12`, `ome12_15`, `ome15_18`, `ome18_21`, `ome21_24`), 
               names_to = "Time")

# Reorder Time Epochs
opioid_long <- opioid_long %>% 
  mutate(Time = factor(Time, levels = c("ome0_3", "ome3_6", "ome6_9", "ome9_12", "ome12_15", "ome15_18", "ome18_21", "ome21_24")))

# Plot Geom_point with median and IQR
ome_time_mediqr <- ggplot(data = opioid_long, aes(x=Time,
                                                  y = value))+
  geom_pointrange(mapping = aes(x = Time, y = value),
                  stat = "summary",
                  fun.min = function(z) {quantile(z,0.25)},
                  fun.max = function(z) {quantile(z,0.75)},
                  fun = median)+
  theme_bw()+
  labs(x="Time", y="Opioid Dose (mg OME)", title="Daily Opioid Dose Over Time")

# Plot Geom_point with median
ome_time_median <- opioid_long %>% 
  group_by(Time) %>% 
  mutate(median = median(value)) %>% 
  ungroup() %>% 
  ggplot()+
  geom_point(aes(x=Time, y = median))+
  geom_line(aes(x=Time, y = median))+
  theme_bw()+
  labs(x="Time", y="Opioid Dose (mg OME)", title="Daily Opioid Dose Over Time")

# Plot with Mean and Error Bars
ome_time_meanerr <- opioid_long %>% 
  group_by(Time) %>% 
  mutate(mean = mean(value)) %>% 
  ungroup() %>% 
  ggplot()+
  geom_point(aes(x=Time, y = mean))+
  geom_line(aes(x=Time, y = mean))+
  theme_bw()+
  labs(x="Time", y="Opioid Dose (mg OME)", title="Daily Opioid Dose Over Time")
  

# Create Daily Opioid Dose (OME)
opioid <- opioid %>% 
  rowwise() %>% 
  mutate(
    morphinfome = (sum(c_across(morph0_3:morph21_24), na.rm = F)*3),
    fentinfome = (sum(c_across(fent0_3:fent21_24), na.rm = F)*2),
    hydroinfome = (sum(c_across(hydro0_3:hydro21_24), na.rm = F)*3)
  ) %>% 
  ungroup()

# Create Total Daily OME
opioid <- opioid %>% 
  mutate(
    total_ome = (morphinfome + fentinfome + hydroinfome)
  )

# Graph Days of Admission with OME Dose
ome_daysofadm <- opioid %>% 
  group_by(da_yofadmission) %>% 
  mutate(mean = mean(total_ome)) %>% 
  ungroup() %>% 
  ggplot()+
  geom_line(aes(x=da_yofadmission, y = mean))+
  theme_bw()+
  labs(x="Days of Admission", y="Opioid Dose (mg OME)", title="Daily Opioid Dose Over Time")+
  coord_cartesian(x = c(0, 7))

hist(opioid$total_ome)

# Replace 0 values with the average between the 2 nonzero numbers (as doses may be validated a couple of hours after they are given)
#opioid <- opioid %>% 
  #replace(.=="NULL", NA) %>% 
  #ifelse(is.na(x) | is.na(c(0, head(x,-1))), 
       #with(rle(na.locf(x, fromLast=T)), rep(values/lengths, lengths)), x)


```

